Understanding the Impact of Mortgage State and Income on Household Poverty Status

### 2024-04-19 paper/US_Poverty_Analysis_2019.qmd
final checks

# read in raw data
raw_poverty_data <-  read_csv(here::here("data/raw_data/raw_poverty_data.csv"))
```

@@

## Model justification
Regarding the relationship between mortgage state and poverty state, we anticipate that both owners without mortgages and owners with mortgages are less likely to experience poverty. Owners without mortgages typically indicate financial stability, as they have already paid off their homes and can allocate financial resources toward household necessities, other than paying back the mortgage / rent. Also, without monthly mortgage or rent payments, their financial burden is reduced, further decreasing the likelihood of poverty. Similarly, owners with mortgages demonstrate financial responsibility by securing loans from banks. The approval of these loans suggests that the bank is confident in owner's ability to manage repayments, indicating a lower possibility of poverty. On the other hand, renters are more likely to experience poverty due to their lack of property ownership and the hidden implication that they do not earn enough to purchase their own property. Renters typically have fewer savings and may earn insufficient salaries to cover rental and essential expenses. Consequently, the combination of rental payments and other financial obligations increases their vulnerability to poverty.

In terms of income levels, we expect a positive relationship between household's income level and their poverty status. This expectation arises from common sense that higher income typically reduces the likelihood of experiencing poverty. If a family is earning a lot of money, then naturally it would not be in poverty. Households that are earning less than 5k annually in the United States obviously have a larger chance in being in poverty due to the high cost of living. Their income probably barely covers or cannot cover for essential expenses such as housing, transportation, or food.

@@

## Relationship betwen Income and Poverty Status
The relationship between a household's income after tax and its poverty status is characterized by a negative correlation, wherein higher income corresponds to a reduced probability of experiencing poverty. This correlation shows the role of financial resources in mitigating economic hardship and facilitating socio-economic well-being. However, in the United States, the cost of living is notably high, low-income households often struggle to meet basic needs, resulting in an unending cycle of poverty. About half of lower-income parents (52%) say they have not had enough money for food or their rent or mortgage. Some 45% of employed parents with lower incomes say they would be very worried about losing pay if they needed to take a day or two off from work to deal with a sick child or because of child care issues [@Braga_2022]. This type of delimma for low-income households makes them harder to pull themselves out of poverty.

In 2019, the poverty guideline for a one-person household was set at $12,490 annually [@ASPE]. So, it is evident that families earning less than $5,000 are definitely living below the poverty line. This is supported by my model result. Conversely, individuals with high incomes are less likely to experience poverty, as their financial resources enable them to comfortably cover living expenses. They will be able to have savings accumulation, further consolidating their financial security and contributing to wealth accumulation over time.

## US Government Programs to Help
For household in poverty, escaping poverty is tough, but the US government and communities are stepping up to help. For example, single parents often struggle with low incomes. But they can get help from government programs such as the Supplemental Nutrition Assistance Program (SNAP) [@snap], commonly known as food stamps, which provides eligible low-income individuals and families with funds to purchase nutritious food. They can also receive child support payments to help cover expenses for their kids. Families with low incomes and lots of kids can also get help. Programs like the Child Tax Credit (CTC) [@ctc], which provides tax credits to eligible families with qualifying children. Other programs, like housing assistance and Medicaid, provide support with things like affordable housing and healthcare. These programs are all aimed at helping families in need to get back on their feet and improve their poverty situation.

>>>>

# read in raw data
raw_poverty_data <-  read_parquet(here::here("data/raw_data/raw_poverty_data.parquet"))
```

@@

## Model justification
Regarding the relationship between mortgage state and poverty state, we anticipate that both owners without mortgages and owners with mortgages are less likely to experience poverty. Owners without mortgages typically indicate financial stability, as they have already paid off their homes and can allocate financial resources toward household necessities. Also, without monthly mortgage or rent payments, their financial burden is reduced, further decreasing the likelihood of poverty. Similarly, owners with mortgages demonstrate financial responsibility by securing loans from banks. The approval of these loans suggests that the bank is confident in owner's ability to manage repayments, indicating a lower possibility of poverty. On the other hand, renters are more likely to experience poverty due to their lack of property ownership and the hidden implication that they do not earn enough to purchase their own property. Renters typically have fewer savings and may earn insufficient salaries to cover rental and essential expenses. Consequently, the combination of rental payments and other financial obligations increases their vulnerability to poverty.

In terms of income levels, we expect a negative relationship between household's income level and their likelihood to be in poverty. This expectation arises from our common sense that higher income typically reduces the likelihood of experiencing poverty. If a family is earning a lot of money, then naturally it would not be in poverty. Households that are earning less than 5k annually in the United States obviously have a larger chance in being in poverty due to the high cost of living. Their income probably barely covers or cannot cover for essential expenses such as housing, transportation, or food.

@@

## Relationship betwen Income and Poverty Status
The relationship between a household's income after tax and its poverty status is characterized by a negative correlation, wherein higher income corresponds to a reduced probability of experiencing poverty. This correlation shows the role of financial resources in mitigating economic hardship and facilitating socio-economic well-being. In 2019, the poverty guideline for a one-person household was set at $12,490 annually [@ASPE]. So, it is evident that families earning less than $5,000 are definitely living below the poverty line. This is supported by my model result. Individuals with high incomes are less likely to experience poverty, as their financial resources enable them to comfortably cover living expenses. They will be able to have savings accumulation, further consolidating their financial security and contributing to wealth accumulation over time.

In the United States, the cost of living is notably high, low-income households often struggle to meet basic needs, resulting in an unending cycle of poverty. About half of lower-income parents (52%) say they have not had enough money for food or their rent or mortgage. 45% of employed parents with lower incomes say they would be very worried about losing pay if they needed to take a day or two off from work to deal with a sick child or because of child care issues [@Braga_2022]. This type of delimma for low-income households makes them harder to pull themselves out of poverty.

## US Government Programs to Help
For household in poverty, escaping poverty is tough, but the US government and communities are stepping up to help. For example, single parents often struggle with low incomes. But they can get help from government programs such as the Supplemental Nutrition Assistance Program (SNAP) [@snap], commonly known as food stamps, which provides eligible low-income individuals and families with funds to purchase nutritious food. They can also receive child support payments to help cover expenses for their kids. Families with low incomes and lots of kids can also get help. Programs like the Child Tax Credit (CTC) [@ctc], which provides tax credits to eligible families with qualifying children. The credit currently provides up to $2,000 per child to about 40 million families every year []. So obviously, these programs are a huge help for low-income family with kids. Other programs, like housing assistance and Medicaid, provide support with things like affordable housing and healthcare. These programs are all aimed at helping families in need to get back on their feet and improve their poverty situation.
